---
title: "EDA lesson notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lesson 4, Problem Set: Explore One Variable

Start by loading necessary modules and the diamonds data set from ggplot
package.

```{r}

library(ggplot2)
data(diamonds)
summary(diamonds)

```

More info about the diamonds data set are available as well.

```{r}

?diamonds

```

We can see the number of observations and the number of variables by using the
`dim()` function.

```{r}

dim(diamonds)

```

Variables in the data set can be accessed using the `$` operator. Accessing
categorical variables this way also gives a short explanation about the
variable levels.

```{r}

diamonds$cut

```

To check how many diamonds there are in different price classes use vector
subsetting and the `len()` function.

```{r}

cheap_diamonds <- diamonds$price[diamonds$price < 500]
length(cheap_diamonds)

very_cheap_diamonds <- diamonds$price[diamonds$price < 250]
length(very_cheap_diamonds)

expensive_diamonds <- diamonds$price[diamonds$price > 15000]
length(expensive_diamonds)

```

Now for some graphics.

Create a histogram of the price of all the diamonds in the diamond data set.
Also write out summary statistics for good measure.

```{r}

qplot(price, data = diamonds, geom = "histogram")
summary(diamonds$price)

```

Explore the largest peak in the
price histogram created earlier.

Try limiting the x-axis, altering the bin width,
and setting different breaks on the x-axis.

```{r}

qplot(price, data = diamonds, geom = "histogram", bins = 50) +
scale_x_continuous(limits = c(300,1500), breaks = seq(300,1500,100)) 

```

Break out the histogram of diamond prices by cut. Use the `scales = "free"`
parameter to avoid forcing the same Y axis for all plots.

```{r}

qplot(x = price, data = diamonds, geom = "histogram", bins = 50) +
facet_wrap(~cut, ncol = 3, scales = "free")
by(diamonds$price, diamonds$cut, summary)

```

Create a histogram of price per carat and facet it by cut. Make adjustments to
the code from the previous exercise to get started. (Changing to using qqplot
from this exercise onwards.)

```{r}

ggplot(aes(x = price/carat), data = diamonds) + 
  geom_histogram(binwidth = 0.03) + 
  scale_x_log10() + 
  facet_wrap(~cut, scales = "free")

```

Investigate the price of diamonds using box plots,
numerical summaries, and the color variable.

```{r}

ggplot(aes(x = color, y = price), data = diamonds) + 
    geom_boxplot()

```

Use the `IQR()` function and find the different interquartile ranges.

```{r}

# Creating vector with prices for "D" diamonds
diamond_price_d <- subset(diamonds, color == "D")$price

# Creating vector with prices for "J" diamonds
diamond_price_j <- subset(diamonds, color == "J")$price

# Calculating first and third quartile by splitting vector in middle and taking the median.
median_d <- median(diamond_price_d)
median(subset(diamond_price_d, diamond_price_d <= median_d))
median(subset(diamond_price_d, diamond_price_d > median_d))

# Calculating first and third quartile by splitting vector in middle and taking the median.
median_j <- median(diamond_price_j)
median(subset(diamond_price_j, diamond_price_j <= median_j))
median(subset(diamond_price_j, diamond_price_j > median_j))

# IQR for both types of diamonds
IQR(diamond_price_d)
IQR(diamond_price_j)

```

Using boxplot to plot the relationship between price per carat and diamond color.

```{r}

qplot(color, price/carat, data = diamonds, geom = "boxplot")

```

Now we will be using a frequency polygon to investigate the diamonds carat variable.

```{r}

qplot(data = diamonds, x = carat, geom = "freqpoly", binwidth = 0.05) + 
  scale_x_continuous(breaks=seq(0,5,0.2))

```


Loading a dataset from gapminder containing unemployment data from almost 200 different countries under more than 15 years. Skipping the last row in the file since it is empty.

```{r}

library("readr")

# Setting explicit columns since last column is not correctly formatted.
# Likewise only reading in the first 178 rows since last row is empty.
employment_data <- read_csv("employment_above_15.csv",
                            col_names = c("country",
                                          "1991",
                                          "1992",
                                          "1993",
                                          "1994",
                                          "1995",
                                          "1996",
                                          "1997",
                                          "1998",
                                          "1999",
                                          "2000",
                                          "2001",
                                          "2002",
                                          "2003",
                                          "2004",
                                          "2005",
                                          "2006",
                                          "2007",
                                          "NA"),
                            col_types = cols("country" = col_character(),
                                             "1991" = col_double(),
                                             "1992" = col_double(),
                                             "1993" = col_double(),
                                             "1994" = col_double(),
                                             "1995" = col_double(),
                                             "1996" = col_double(),
                                             "1997" = col_double(),
                                             "1998" = col_double(),
                                             "1999" = col_double(),
                                             "2000" = col_double(),
                                             "2001" = col_double(),
                                             "2002" = col_double(),
                                             "2003" = col_double(),
                                             "2004" = col_double(),
                                             "2005" = col_double(),
                                             "2006" = col_double(),
                                             "2007" = col_double(),
                                             "NA" = col_skip()),
                            skip = 1,
                            n_max = 178)

employment_data

```

Tyding up the data by adding years as a variable.

```{r}

library(tidyr)
employment_data <- gather(employment_data, 
                          "1991",    
                          "1992",    
                          "1993",    
                          "1994",    
                          "1995",    
                          "1996",    
                          "1997",    
                          "1998",    
                          "1999",    
                          "2000",    
                          "2001",    
                          "2002",    
                          "2003",    
                          "2004",    
                          "2005",    
                          "2006",    
                          "2007",    
                          key = "year", 
                          value = "employment_rate")
head(employment_data)
```

Print out countries with the highest average employment in the data set.

```{r}

library(dplyr)

avg_employment_rate <- employment_data %>%
  group_by(country) %>%
  summarise(average_employment = mean(employment_rate))

top_n(avg_employment_rate, 5, average_employment)

```
... and the lowest.

```{r}

top_n(avg_employment_rate, -5, average_employment)

```

Plot line graph of all countries data.

```{r}
# ggplot(data=melted, aes(x=time, y=value, group=variable)) + geom_line()

ggplot(data = employment_data, 
       aes(x = year,  y = employment_rate, group = country)) + 
  geom_line()
  
```

A bit too messy let's instead plot the top five countries 1991 and the top five countries 2007 over the total time of the data set timeline.

```{r}

top_5_1991 <- employment_data %>% 
  filter(year == "1991") %>% 
  arrange(employment_rate) %>% 
  top_n(5)

top_5_2007 <- employment_data %>% 
  filter(year == "2007") %>% 
  arrange(employment_rate) %>% 
  top_n(5)

top_countries <- union(top_5_1991, top_5_2007)["country"]
employment_data_of_interest <- inner_join(top_countries, employment_data)

ggplot(data = employment_data_of_interest, 
       aes(x = year,  y = employment_rate, group = country, color = country)) + 
  geom_line()

```

Creating a new data frame containing one variable which is the employment rate for people between 15 and 24 specifically.

```{r}

employment_data_15_to_24 <- read_csv("employment_15_to_24.csv",
                                     col_names = c("country",
                                                   "1991",
                                                   "1992",
                                                   "1993",
                                                   "1994",
                                                   "1995",
                                                   "1996",
                                                   "1997",
                                                   "1998",
                                                   "1999",
                                                   "2000",
                                                   "2001",
                                                   "2002",
                                                   "2003",
                                                   "2004",
                                                   "2005",
                                                   "2006",
                                                   "2007",
                                                   "NA"),
                                     col_types = cols("country" = col_character(),
                                                      "1991" = col_double(),
                                                      "1992" = col_double(),
                                                      "1993" = col_double(),
                                                      "1994" = col_double(),
                                                      "1995" = col_double(),
                                                      "1996" = col_double(),
                                                      "1997" = col_double(),
                                                      "1998" = col_double(),
                                                      "1999" = col_double(),
                                                      "2000" = col_double(),
                                                      "2001" = col_double(),
                                                      "2002" = col_double(),
                                                      "2003" = col_double(),
                                                      "2004" = col_double(),
                                                      "2005" = col_double(),
                                                      "2006" = col_double(),
                                                      "2007" = col_double(),
                                                      "NA" = col_skip()),
                                     skip = 1,
                                     n_max = 178)

employment_data_15_to_24 <- gather(employment_data_15_to_24, 
                                   "1991",    
                                   "1992",    
                                   "1993",    
                                   "1994",    
                                   "1995",    
                                   "1996",    
                                   "1997",    
                                   "1998",    
                                   "1999",    
                                   "2000",    
                                   "2001",    
                                   "2002",    
                                   "2003",    
                                   "2004",    
                                   "2005",    
                                   "2006",    
                                   "2007",    
                                   key = "year", 
                                   value = "employment_rate")

employment_data_15_to_24_of_interest <- inner_join(top_countries, 
                                                   employment_data_15_to_24)

# join together the new and old dataframes.

employment_data_of_interest <- left_join(employment_data_of_interest,
                                         employment_data_15_to_24_of_interest, 
                                         by = c("country", "year"),
                                         suffix = c("_all", "_15_to_24"))

employment_data_of_interest <- gather(employment_data_of_interest,
                                      "employment_rate_all",
                                      "employment_rate_15_to_24",
                                      key = "group",
                                      value = "employment_rate")

```

Plotting the result against the employment rate for all ages.

```{r}

ggplot(data = employment_data_of_interest, 
       aes(x = year,  y = employment_rate, group = country, color = country)) + 
  geom_line() +
  facet_grid(group ~ .)

```

Now let's check out the birthdays of our imaginary facebook friends.
Start by reading in the data.

```{r}

library(readr)

birthdays <- read_csv("birthdaysExample.csv")
birthdays$dates <- as.Date(parse_date(birthdays$dates, "%m/%d/%y"))

```

Let's answer the following questions.

* How many people share your birthday? Do you know them?
  (Reserve time with them or save money to buy them a gift!)

```{r}

subset(birthdays, dates == dmy("030514"))

```

Actually no one does share my birthday...


* Which month contains the most number of birthdays?
* How many birthdays are in each month?

```{r}

qplot(x = month(dates), data = birthdays, geom = "histogram", binwidth = 1) + 
  scale_x_continuous(breaks = seq(0, 12, 1)) +
  stat_bin(binwidth = 1, geom='text', aes(label= ..count.. ) , color="white" ,vjust = 1.5) +
  labs(y = "Number of birthdays", x = "Month")

```

March seems to be the winner with 98 birthdays.


* Which day of the year has the most number of birthdays?

```{r}

top_birthday <- birthdays %>% 
  group_by(days = yday(birthdays$dates)) %>% 
  count(days) %>% 
  ungroup() %>% 
  top_n(1)

as.Date(top_birthday$days - 1, origin = "2014-01-01")
```

It is a tie between the above three.


* Do you have at least 365 friends that have birthdays on everyday
  of the year?

```{r}

length(unique(birthdays$dates))

```

We only have 348 unique values in our data set meaning that there aren't enought unique birthdays to fill up with at least one per day.

