Prosper loan data analysis by Eric Persson
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse)
library(gridExtra)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data by reading just the columns we would like to use and giving them
# a suitable data type.
loan_data <- read_csv("data/prosperLoanData.csv",
                      col_types = cols_only(
                        "ListingCreationDate" = col_datetime(),
                        "Term" = col_integer(),
                        "LoanStatus" = col_character(),
                        "ClosedDate" = col_datetime(),
                        "ListingCategory (numeric)" = col_integer(),
                        "Occupation" = col_character(),
                        "EmploymentStatus" = col_character(),
                        "EmploymentStatusDuration" =  col_integer(),
                        "IsBorrowerHomeowner" = col_logical(),
                        "CurrentlyInGroup" = col_logical(),
                        "DebtToIncomeRatio" = col_double(),
                        "IncomeRange" = col_character(),
                        "TotalProsperLoans" = col_integer(),
                        "LoanOriginalAmount" = col_integer(),
                        "LoanOriginationDate" = col_datetime(), 
                        "MonthlyLoanPayment" = col_double(),
                        "BorrowerRate" = col_double(),
                        "InvestmentFromFriendsCount" = col_integer(),
                        "InvestmentFromFriendsAmount" = col_double(),
                        "Investors" = col_integer()
                      ))

# Replace NA values in the EmploymentStatusDuration and TotalProsperLoans
# variables with zeros. NA values in EmploymentStatus and Occupation replaced
# with counterparts already exsisting in the data.
loan_data <- loan_data %>% 
  replace_na(list(EmploymentStatusDuration = 0,
                  TotalProsperLoans = 0,
                  EmploymentStatus = "Not available",
                  Occupation = "Unknown"))

# Translating the numeric ListingCategory values to readable string data.
listingCategory <- tribble(~"ListingCategory (numeric)", ~"ListingCategory",
                           0, "Not available", 
                           1, "Debt Consolidation", 
                           2, "Home Improvement", 
                           3, "Business", 
                           4, "Personal Loan", 
                           5, "Student Use", 
                           6, "Auto", 
                           7, "Other", 
                           8, "Baby&Adoption", 
                           9, "Boat", 
                           10, "Cosmetic Procedure", 
                           11, "Engagement Ring", 
                           12, "Green Loans", 
                           13, "Household Expenses", 
                           14, "Large Purchases", 
                           15, "Medical/Dental", 
                           16, "Motorcycle", 
                           17, "RV", 
                           18, "Taxes", 
                           19, "Vacation", 
                           20, "Wedding Loans")
loan_data <- loan_data %>% left_join(listingCategory, by = "ListingCategory (numeric)")
loan_data["ListingCategory (numeric)"] <- NULL

```

For this exploratory data analysis we are having a look at loan listings data 
from a web service called (Prosper)[https://www.prosper.com] to try to figure
out who are using the service, why they are taking a loan, and what eventually
happends to that loan.

Since the original dataset contains over 80 variables I have picked out a subset
which we will use for our analysis based on the above stated questions.
Initially some light data wrangling was also made to either make the data set
more readable and to handle NA values.

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

Let's start by having a look at the summary statistics for the data to see what
we have to work with.

```{r echo=FALSE, Univariate_Plots_1}

summary(loan_data)

```

Based on the above numerical data our typical loan taker is a first time prosper
user with equal posibility to be a homeowner or not, taking a loan over 36 to 40
months with an interest rate of around 19%. The typical size of a loan is $6500.

Next we plot the numbers of occurrencies for the nominal variables in our data set.

```{r echo=FALSE, warning=FALSE, fig.align="center", fig.height=3, fig.width=9, Univariate_Plots_2}

# Create a dataframe with the counts for all employment statuses and plot the
# results in a tilted bar plot.
ranked_employment_statuses <- loan_data %>% 
  group_by(EmploymentStatus) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_employment_statuses) +
  geom_col(aes(reorder(EmploymentStatus, Count), Count)) +
  labs(title = "Ranked Employment Statuses",
       x = "Employment Status ordered by Count") +
  coord_flip()

# Create dataframe with counts for each type of occupation in the data set,
# arrange it by counts, and then select the top ten occupations. The data is
# then plotted in a tilted bar plot.
top_10_occupations <- loan_data %>% 
  group_by(Occupation) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  top_n(10, Count)
ggplot(top_10_occupations) +
  geom_col(aes(reorder(Occupation, Count), Count)) +
  labs(title = "Top 10 Occupations",
       x = "Occupation ordered by Count") +
  coord_flip()

# Create a dataframe with counts for each different type of loan status. Plot
# the resulting dataframe according to the order given by the limits parameter
# in the scale_x_limits layer.
loan_statuses <- loan_data %>% 
  group_by(LoanStatus) %>% 
  summarise(Count = n())
ggplot(loan_statuses) +
  geom_col(aes(LoanStatus, Count)) +
  labs(title = "Loan Statuses") +
  scale_x_discrete(limits = c("Chargedoff",
                              "Defaulted",
                              "Past Due (>120 days)",
                              "Past Due (91-120 days)",
                              "Past Due (61-90 days)",
                              "Past Due (31-60 days)",
                              "Past Due (16-30 days)",
                              "Past Due (1-15 days)",
                              "Cancelled",
                              "Completed",
                              "Current")) +
  coord_flip()

# Create and plot a dataframe with the counts for all the exsisting income
# ranges in the data set. The labels are tilted by 45 degree for readability.
count_income_ranges <- loan_data %>% 
  group_by(IncomeRange) %>% 
  summarise(Count = n())
ggplot(count_income_ranges) +
  geom_col(aes(IncomeRange, Count)) +
  labs(title = "Income Ranges") +
  theme(axis.text.x=element_text(angle=20, vjust=0.5))

# Create and plot a dataframe with the counts for all exsisting listing
# categories in the data set.
ranked_listing_categories <- loan_data %>% 
  group_by(ListingCategory) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_listing_categories) +
  geom_col(aes(reorder(ListingCategory, Count), Count)) +
  labs(title = "Ranked Listing Categories",
       x = "Listing Categories ordered by count") +
  coord_flip()

```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

From the above plots we can see that most of the loan takers are employed but
the type of occupation is seldom given with the vague "Professional" and "Other"
occupation types both being in the top ten. A majority of loans are still being
repaid but there are also an substantial amount of past due, defaulted or
chargedoff loans.

Further, the income range looks to be fairly normalised with an expected value
somewhere around $50,000. Lastly, we have the listing categories for the loan
listings and we can see that roughly half of the the reasons given for the loans
through prosper is debt consolidation followed by the rather vague "Not
Available" and "Other" categories in the top three.

Now, after have gone through and had an initial look at all the variables in the
data set let's revisit and plot some of the more interesting numerical variables
to see how they are distributed.

```{r echo=FALSE, warning=FALSE, fig.align="center", fig.height=3, fig.width=9, Univariate_Plots_3}

# Area plot for loan origination and closing dates. Binned at 1 month periods
# for 100 months for the period where we have available data for the whole
# month.
loan_dates <- loan_data %>% 
  select(one_of("LoanOriginationDate", "ClosedDate")) %>% 
  gather("LoanOriginationDate", 
         "ClosedDate", 
         key = "Type", 
         value = "Date",
         factor_key = TRUE)
ggplot(subset(loan_dates, Date < as.Date("2014-03-01") & Date > as.Date("2005-12-01"))) +
  geom_histogram(aes(x = Date, fill = Type), 
            alpha = 0.5, 
            stat = "bin", 
            bins = 100,
            position = "identity") +
  theme(legend.position = "bottom")

# Plotting original loan amounts binned by $1000
ggplot(loan_data) +
  geom_histogram(aes(LoanOriginalAmount), bins = 35) +
  scale_x_continuous(breaks = seq(0, 35000, 5000))

# Loan term lenghts 
ggplot(loan_data) +
  geom_histogram(aes(Term), bins = 5) +
  scale_x_continuous(breaks = seq(0, 60, 12))

# Monthly Loan Payments
ggplot(loan_data) +
  geom_histogram(aes(MonthlyLoanPayment), bins = 46) +
  scale_x_continuous(breaks = seq(0, 2300, 100))

# $DebtToIncomeRatio using log10 scale for y to compensate for the number of
# values around 0.25
ggplot(loan_data) +
  geom_histogram(aes(DebtToIncomeRatio), bins = 41) +
  scale_x_continuous(breaks = seq(0, 10, 0.5)) +
  scale_y_log10()

```

By plotting the above values we discovered some interesting facts such as that
the loan term probably are locked at one, three or five years. We also saw that
the usual amounts being borrowed are grouped around even $5000 numbers with a
maximum at $35000. Lastly, the loan origination dates clearly shows effects from
the 2008 recession and also an peak in new loans later years which we still
aren't able to explain. Let's proceed with some further analysis before
investigating inter-variable relations in the bivariate section.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!