Prosper loan data analysis by Eric Persson
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse)
library(gridExtra)
library(lubridate)
library(GGally)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data by reading just the columns we would like to use and giving them
# a suitable data type.
loan_data <- read_csv("data/prosperLoanData.csv",
                      col_types = cols_only(
                        "ListingCreationDate" = col_datetime(),
                        "Term" = col_integer(),
                        "LoanStatus" = col_character(),
                        "ClosedDate" = col_datetime(),
                        "ListingCategory (numeric)" = col_integer(),
                        "Occupation" = col_character(),
                        "EmploymentStatus" = col_character(),
                        "EmploymentStatusDuration" =  col_integer(),
                        "IsBorrowerHomeowner" = col_logical(),
                        "CurrentlyInGroup" = col_logical(),
                        "DebtToIncomeRatio" = col_double(),
                        "IncomeRange" = col_character(),
                        "TotalProsperLoans" = col_integer(),
                        "LoanOriginalAmount" = col_integer(),
                        "LoanOriginationDate" = col_datetime(), 
                        "MonthlyLoanPayment" = col_double(),
                        "BorrowerRate" = col_double(),
                        "InvestmentFromFriendsCount" = col_integer(),
                        "InvestmentFromFriendsAmount" = col_double(),
                        "Investors" = col_integer()
                      ))

# Replace NA values in the EmploymentStatusDuration and TotalProsperLoans
# variables with zeros. NA values in EmploymentStatus and Occupation replaced
# with counterparts already exsisting in the data.
loan_data <- loan_data %>% 
  replace_na(list(EmploymentStatusDuration = 0,
                  TotalProsperLoans = 0,
                  EmploymentStatus = "Not available",
                  Occupation = "Unknown"))

# Translating the numeric ListingCategory values to readable string data.
listingCategory <- tribble(~"ListingCategory (numeric)", ~"ListingCategory",
                           0, "Not available", 
                           1, "Debt Consolidation", 
                           2, "Home Improvement", 
                           3, "Business", 
                           4, "Personal Loan", 
                           5, "Student Use", 
                           6, "Auto", 
                           7, "Other", 
                           8, "Baby&Adoption", 
                           9, "Boat", 
                           10, "Cosmetic Procedure", 
                           11, "Engagement Ring", 
                           12, "Green Loans", 
                           13, "Household Expenses", 
                           14, "Large Purchases", 
                           15, "Medical/Dental", 
                           16, "Motorcycle", 
                           17, "RV", 
                           18, "Taxes", 
                           19, "Vacation", 
                           20, "Wedding Loans")
loan_data <- loan_data %>% left_join(listingCategory, by = "ListingCategory (numeric)")
loan_data["ListingCategory (numeric)"] <- NULL

# Set the proper order for factors
loan_data$LoanStatus <- factor(loan_data$LoanStatus, 
                               levels = c("Chargedoff",
                                          "Defaulted",
                                          "Past Due (>120 days)",
                                          "Past Due (91-120 days)",
                                          "Past Due (61-90 days)",
                                          "Past Due (31-60 days)",
                                          "Past Due (16-30 days)",
                                          "Past Due (1-15 days)",
                                          "Cancelled",
                                          "Completed",
                                          "Current"))

loan_data$IncomeRange <- factor(loan_data$IncomeRange, 
                               levels = c("Not displayed",
                                          "Not employed",
                                          "$0",
                                          "$1-24,999",
                                          "$25,000-49,999",
                                          "$50,000-74,999",
                                          "$75,000-99,999",
                                          "$100,000+"))
```

For this exploratory data analysis we are having a look at loan listings data 
from a web service called [Prosper](https://www.prosper.com) to try to figure
out who are using the service, why they are taking a loan, and what eventually
happens to that loan.

Since the original data set contains over 80 variables I have picked out a subset
which we will use for our analysis based on the above stated questions.
Initially some light data wrangling was also made to either make the data set
more readable and to handle NA values.

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your data set. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your data set. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

Let's start by having a look at the summary statistics for the data to see what
we have to work with.

```{r echo=FALSE, Univariate_Plots_1}

summary(loan_data)

```

Based on the above numerical data our typical loan taker is a first time prosper
user with equal possibility to be a homeowner or not, taking a loan over 36 to 40
months with an interest rate of around 19%. The typical size of a loan is $6500.

Next we plot the numbers of occurrences for the nominal variables in our data set.

```{r echo=FALSE, warning=FALSE, fig.align="center", Univariate_Plots_2}

# Create a dataframe with the counts for all employment statuses and plot the
# results in a tilted bar plot.
ranked_employment_statuses <- loan_data %>% 
  group_by(EmploymentStatus) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_employment_statuses) +
  geom_col(aes(reorder(EmploymentStatus, Count), Count)) +
  labs(title = "Ranked Employment Statuses",
       x = "Employment Status ordered by Count") +
  coord_flip()

# Create dataframe with counts for each type of occupation in the data set,
# arrange it by counts, and then select the top ten occupations. The data is
# then plotted in a tilted bar plot.
top_10_occupations <- loan_data %>% 
  group_by(Occupation) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  top_n(10, Count)
ggplot(top_10_occupations) +
  geom_col(aes(reorder(Occupation, Count), Count)) +
  labs(title = "Top 10 Occupations",
       x = "Occupation ordered by Count") +
  coord_flip()

# Create a dataframe with counts for each different type of loan status. Plot
# the resulting dataframe according to the order given by the limits parameter
# in the scale_x_limits layer.
# series$type2 <- factor(series$type, levels = c('c', 'b', 'd', 'a'))
loan_statuses <- loan_data %>% 
  group_by(LoanStatus) %>% 
  summarise(Count = n())
ggplot(loan_statuses) +
  geom_col(aes(LoanStatus, Count)) +
  labs(title = "Loan Statuses") +
  coord_flip()

# Create and plot a dataframe with the counts for all the exsisting income
# ranges in the data set. The labels are tilted by 45 degree for readability.
count_income_ranges <- loan_data %>% 
  group_by(IncomeRange) %>% 
  summarise(Count = n())
ggplot(count_income_ranges) +
  geom_col(aes(IncomeRange, Count)) +
  labs(title = "Income Ranges") +
  theme(axis.text.x=element_text(angle=20, vjust=0.5)) +
  coord_flip()

# Create and plot a dataframe with the counts for all exsisting listing
# categories in the data set.
ranked_listing_categories <- loan_data %>% 
  group_by(ListingCategory) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_listing_categories) +
  geom_col(aes(reorder(ListingCategory, Count), Count)) +
  labs(title = "Ranked Listing Categories",
       x = "Listing Categories ordered by count") +
  coord_flip()

```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

From the above plots we can see that most of the loan takers are employed but
the type of occupation is seldom given with the vague "Professional" and "Other"
occupation types both being in the top ten. A majority of loans are still being
repaid but there are also an substantial amount of past due, defaulted or
charged-off loans.

Further, the income range looks to be fairly normalized with an expected value
somewhere around $50,000. Lastly, we have the listing categories for the loan
listings and we can see that roughly half of the the reasons given for the loans
through prosper is debt consolidation followed by the rather vague "Not
Available" and "Other" categories in the top three.

Now, after have gone through and had an initial look at all the variables in the
data set let's revisit and plot some of the more interesting numerical variables
to see how they are distributed over time.

```{r echo=FALSE, warning=FALSE, fig.align="center", Univariate_Plots_3}

# Bin plots for loan origination and closing dates. Binned at 1 month periods
# for 100 months for the period where we have available data for the whole
# month. We also calculate the cumulative sum of Originations and Closing per
# entry to get the amount of active listings for the period 2005-12-01 to 2014-03-01.
ggplot(subset(loan_data, 
              LoanOriginationDate < as.Date("2014-03-01") & 
                LoanOriginationDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(LoanOriginationDate), stat = "bin", bins = 100) +
  scale_y_continuous(limits = c(0, 6000),
                     breaks = seq(0, 6000, 1000))

ggplot(subset(loan_data, 
              ClosedDate < as.Date("2014-03-01") & 
                ClosedDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(ClosedDate), stat = "bin", bins = 100) +
  scale_y_continuous(limits = c(0, 6000),
                     breaks = seq(0, 6000, 1000))

Originations <- loan_data %>%
  group_by(LoanOriginationDate) %>%
  summarise(OriginationCount = n(), 
            MeanBorrowerRate = mean(BorrowerRate))
Closings <- loan_data %>% 
  group_by(ClosedDate) %>% 
  summarise(ClosedCount = -n())
Delta <- tibble(Date = seq(ymd_hms("2005-11-15 00:00:00"), 
                           ymd_hms("2014-03-12 00:00:00"), 
                           "days")) %>% 
  left_join(Originations, 
            by = c("Date" = "LoanOriginationDate")) %>% 
  left_join(Closings,
            by = c("Date" = "ClosedDate")) %>% 
  rowwise() %>% 
  mutate(DeltaCount = sum(OriginationCount,
                           ClosedCount,
                           na.rm = TRUE))
ActiveListings <- Delta %>% 
  transform(Count = cumsum(DeltaCount))
ggplot(ActiveListings) + 
  geom_line(aes(Date, Count)) 

# Calculating and plotting monthly active listings changes
MonthlyActiveListings <- ActiveListings %>% 
  group_by(Month = floor_date(Date, "month")) %>% 
  summarise(Delta = sum(DeltaCount))
ggplot(MonthlyActiveListings) + 
  geom_line(aes(Month, Delta))

# Plotting mean borrower rates over time for originating loans
MonthlyOriginations <- Originations %>% 
  group_by(Month = floor_date(LoanOriginationDate, "month")) %>% 
  summarise(MonthlyMeanBorrowerRate = mean(MeanBorrowerRate))
ggplot(MonthlyOriginations) +
  geom_line(aes(Month, MonthlyMeanBorrowerRate))
  

# Plotting original loan amounts binned by $1000
ggplot(loan_data) +
  geom_histogram(aes(LoanOriginalAmount), bins = 35) +
  scale_x_continuous(breaks = seq(0, 35000, 5000))

# Loan term lenghts 
ggplot(loan_data) +
  geom_histogram(aes(Term), bins = 5) +
  scale_x_continuous(breaks = seq(0, 60, 12))

# Monthly Loan Payments
ggplot(loan_data) +
  geom_histogram(aes(MonthlyLoanPayment), bins = 46) +
  scale_x_continuous(breaks = seq(0, 2300, 100)) +
  theme(axis.text.x=element_text(angle=20, vjust=0.5))

# Borrower rates histogram
ggplot(loan_data) +
  geom_histogram(aes(BorrowerRate), bins = 51) +
  scale_x_continuous(breaks = seq(0, 0.5, 0.1))

# $DebtToIncomeRatio using log10 scale for y to compensate for the number of
# values around 0.25
ggplot(loan_data) +
  geom_histogram(aes(DebtToIncomeRatio), bins = 41) +
  scale_x_continuous(breaks = seq(0, 10, 0.5)) +
  scale_y_log10()

```

By plotting the above values we discovered some interesting facts such as that
the loan term probably are artificially locked at one, three or five years. We
also saw that the usual amounts being borrowed are grouped around even $5000
numbers with a maximum at $35000. Lastly, the loan origination dates clearly
shows effects from the 2008 recession and also an peak in new loans later years
which we still aren't able to explain. Let's proceed with some further analysis
before investigating inter-variable relations in the bivariate section.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

The original data set contains more variables than what we could practically
cover in one go so to narrow them down we posed some questions about the data.
Let's revisit these questions to see if we made any discoveries worth noting
already.

First, to see who is using the service we can look at the following variables
mentioned:

* Employment Status
* Occupations
* Income Ranges
* Is Borrower Homeowner
* etc.

Based on the summary data and plots presented, our typical loan taker is
employed, with an unspecified occupation, and probably an income of $25,000 to
$50,000. He/she is currently an homeowner and have debts of about a ratio of
0.22 of their income.

The reason for the loan is most likely debt consolidation with home improvement
and business lying as distant seconds among the specified reasons as seen in the
histogram with ranked listing categories.

To see what eventually happened with the loan we can have a look at the loan
status bin plot giving an overview over the different statuses for all the loans
in the data set. Out of a little over 100,000 loan listings we have a little
over 10,000 that have been defaulted or charged-off(> 150 days overdue with no
reasonable expectation of sufficient payment).

### What is/are the main feature(s) of interest in your dataset?
### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Loan origination dates vs. closed dates together with the data on overdue,
defaulted, or charged-off loans. Using these features together with the above
calculated active loans variable it would be interesting to dive deeper and see
how, and when, the 2008 recession affected the loans taken. Further, since it
seems like the service has enjoyed some explosive growth going in to 2014 it
would be very interesting to see what these added loans are and perhaps why they
have increased.

Another features worth looking in to is the borrower rates which mostly pikes my
interest due to the unclear form of the distribution. Investigating what
variables are correlated and how they affect this blob of values centered
somewhere around 0.2 would be very interesting and perhaps a good candidate for
a regression model and analysis.

### Did you create any new variables from existing variables in the dataset?

Using the Loan origination dates together with the closing dates I calculated a
new variable called active listings to show the volume of current loans on the
service. The calculation where made by taking the difference between originated
and closed loans for each date during a period between 2005 and 2014 and the
calculate the cumulative sum of those differences.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

In addition to make sure that all variables were read in using the correct data
type, I also choose how to handle NA values. For NA values for nominal data was
substituted with an preexisting category being best suited to improve the
readability of following histograms. Numeral values where set to 0 for ordinal
variables where NAs where present this in order to ease calculations in the
analysis while not influencing other statistical measures.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

To continue follow our interesting findings regarding the effects of the 2008
recession on active loan volumes let's plot the change in loan statuses and 
listing categories for the period.

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_1}

# Loan closed statuses for loans over time.
ggplot(subset(loan_data, 
              ClosedDate < as.Date("2014-03-01") & 
                ClosedDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(ClosedDate, fill = LoanStatus), 
                 stat = "bin",
                 bins = 100,
                 position = "fill") +
  scale_fill_manual(values = c("#c7e9c0",
                               "#74c476", 
                               "#238b45",
                               "#005a32"))

# Ratio for listing categories over time.
ggplot(loan_data) +
  geom_histogram(aes(ListingCreationDate, fill = ListingCategory),
                 bins = 100, 
                 position = "fill") +
  scale_fill_manual(values = c("#edf8e9",
                               "#c7e9c0",
                               "#a1d99b",
                               "#74c476", 
                               "#41ab5d",
                               "#238b45",
                               "#005a32",
                               "#eff3ff",
                               "#c6dbef",
                               "#9ecae1",
                               "#6baed6",
                               "#4292c6",
                               "#2171b5",
                               "#084594",
                               "#feedde",
                               "#fdd0a2",
                               "#fdae6b",
                               "#fd8d3c",
                               "#f16913",
                               "#d94801",
                               "#8c2d04"))

```

First, we can see that there have been some change in the gathering of the data
in the shift between 2007 and 2008. Looking at loan statuses we can see that the 
ratio for Chargedoff and Defaulted loans did shift sharply with the amount of 
Chargedoff loans skyrocketing. Analogously, the listing categories went from 
mostly Not available to a mix of all different categories at the same period.

Looking at some of the numerical values in our data set we unfortunately cannot 
see any meaningful correlation between variables, except for the connection 
between monthly payments and the original amount of the loan.

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_2}

# Plotting correlation matrix
ggpairs(subset(loan_data, 
               select = c("BorrowerRate",
                          "LoanOriginalAmount",
                          "MonthlyLoanPayment",
                          "DebtToIncomeRatio",
                          "Investors")),
        lower = list(continuous = wrap("points", 
                                       size = 0.2,
                                       alpha = 0.1))) +
    theme(axis.text.x=element_text(angle=20, vjust=0.5))


```

Following our slight drawback from analysing the continous variables we instead 
focus on the categorical ones. Here we can start to see some slight trends in 
the relationships between all our supporting variables and the borrower rate, 
however not one single variable seems significant enough to prove anything. 

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_3}

# Plotting the term(cut into categories) vs given borrower rate
ggplot(loan_data) +
  geom_boxplot(aes(cut(Term, c(0, 12, 36, 60)),
                 BorrowerRate)) +
  coord_flip()

# Plotting IncomeRange vs borrower rate
ggplot(loan_data) +
  geom_boxplot(aes(IncomeRange,
                 BorrowerRate)) +
  coord_flip()

# Plotting ListingCategory vs borrower rate (Ranked by BorrowerRate median)
ranked_listing_categories_by_borrower_rate <- loan_data %>% 
  group_by(ListingCategory) %>% 
  summarise(Median = median(BorrowerRate)) %>% 
  arrange(Median)
loan_data$ListingCategory <- 
  factor(loan_data$ListingCategory,
         levels = ranked_listing_categories_by_borrower_rate$ListingCategory)
ggplot(loan_data) +
  geom_boxplot(aes(ListingCategory, BorrowerRate)) +
  coord_flip()

# Plotting Occupation vs Borrower rate (Ranked by BorrowerRate median)
ranked_occupation_by_borrower_rate <- loan_data %>% 
  group_by(Occupation) %>% 
  summarise(Median = median(BorrowerRate)) %>% 
  arrange(Median)
loan_data$Occupation <- 
  factor(loan_data$Occupation,
         levels = ranked_occupation_by_borrower_rate$Occupation)
ggplot(loan_data) +
  geom_boxplot(aes(Occupation,
                 BorrowerRate)) +
  coord_flip()

# Plotting HomeOwner vs BorrowerRate
ggplot(loan_data) +
  geom_boxplot(aes(IsBorrowerHomeowner,
                 BorrowerRate)) +
  coord_flip()

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?



### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your data set.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the data set.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!