Prosper loan data analysis by Eric Persson
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse)
library(gridExtra)
library(lubridate)
library(GGally)
library(scales)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data by reading just the columns we would like to use and giving them
# a suitable data type.
loan_data <- read_csv("data/prosperLoanData.csv",
                      col_types = cols_only(
                        "ListingCreationDate" = col_datetime(),
                        "Term" = col_integer(),
                        "LoanStatus" = col_character(),
                        "ClosedDate" = col_datetime(),
                        "ListingCategory (numeric)" = col_integer(),
                        "Occupation" = col_character(),
                        "EmploymentStatus" = col_character(),
                        "EmploymentStatusDuration" =  col_integer(),
                        "IsBorrowerHomeowner" = col_logical(),
                        "CurrentlyInGroup" = col_logical(),
                        "DebtToIncomeRatio" = col_double(),
                        "IncomeRange" = col_character(),
                        "TotalProsperLoans" = col_integer(),
                        "LoanOriginalAmount" = col_integer(),
                        "LoanOriginationDate" = col_datetime(), 
                        "MonthlyLoanPayment" = col_double(),
                        "BorrowerRate" = col_double(),
                        "InvestmentFromFriendsCount" = col_integer(),
                        "InvestmentFromFriendsAmount" = col_double(),
                        "Investors" = col_integer()
                      ))

# Replace NA values in the EmploymentStatusDuration and TotalProsperLoans
# variables with zeros. NA values in EmploymentStatus and Occupation replaced
# with counterparts already exsisting in the data.
loan_data <- loan_data %>% 
  replace_na(list(EmploymentStatusDuration = 0,
                  TotalProsperLoans = 0,
                  EmploymentStatus = "Not available",
                  Occupation = "Unknown"))

# Translating the numeric ListingCategory values to readable string data.
listingCategory <- tribble(~"ListingCategory (numeric)", ~"ListingCategory",
                           0, "Not available", 
                           1, "Debt Consolidation", 
                           2, "Home Improvement", 
                           3, "Business", 
                           4, "Personal Loan", 
                           5, "Student Use", 
                           6, "Auto", 
                           7, "Other", 
                           8, "Baby&Adoption", 
                           9, "Boat", 
                           10, "Cosmetic Procedure", 
                           11, "Engagement Ring", 
                           12, "Green Loans", 
                           13, "Household Expenses", 
                           14, "Large Purchases", 
                           15, "Medical/Dental", 
                           16, "Motorcycle", 
                           17, "RV", 
                           18, "Taxes", 
                           19, "Vacation", 
                           20, "Wedding Loans")
loan_data <- loan_data %>% left_join(listingCategory, by = "ListingCategory (numeric)")
loan_data["ListingCategory (numeric)"] <- NULL

# Set the proper order for factors
loan_data$LoanStatus <- factor(loan_data$LoanStatus, 
                               levels = c("Chargedoff",
                                          "Defaulted",
                                          "Past Due (>120 days)",
                                          "Past Due (91-120 days)",
                                          "Past Due (61-90 days)",
                                          "Past Due (31-60 days)",
                                          "Past Due (16-30 days)",
                                          "Past Due (1-15 days)",
                                          "Cancelled",
                                          "Completed",
                                          "Current"))

loan_data$IncomeRange <- factor(loan_data$IncomeRange, 
                               levels = c("Not displayed",
                                          "Not employed",
                                          "$0",
                                          "$1-24,999",
                                          "$25,000-49,999",
                                          "$50,000-74,999",
                                          "$75,000-99,999",
                                          "$100,000+"))
```

For this exploratory data analysis we are having a look at loan listings data 
from a web service called [Prosper](https://www.prosper.com) to try to figure
out who are using the service, why they are taking a loan, and how much that
loan costs.

Since the original data set contains over 80 variables I have picked out a subset
which we will use for our analysis loosely based on the above stated questions.
Initially some light data wrangling was also made to either make the data set
more readable and to handle NA values.

# Univariate Plots Section

Let's start by having a look at the summary statistics for the data to see what
we have to work with.

```{r echo=FALSE, Univariate_Plots_1}

summary(loan_data)

```

Based on the above numerical data our typical loan taker is a first time prosper
user with equal possibility to be a homeowner as not, taking a loan over 36 to 40
months with an interest rate of around 19%. The typical size of a loan is $6500.

Next we plot the numbers of occurrences for the nominal variables in our data set.

```{r echo=FALSE, warning=FALSE, fig.align="center", Univariate_Plots_2}

# Create a dataframe with the counts for all employment statuses and plot the
# results in a tilted bar plot.
ranked_employment_statuses <- loan_data %>% 
  group_by(EmploymentStatus) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_employment_statuses) +
  geom_col(aes(reorder(EmploymentStatus, Count), Count)) +
  labs(x = "Employment Status ordered by Count") +
  coord_flip()

# Create dataframe with counts for each type of occupation in the data set,
# arrange it by counts, and then select the top ten occupations. The data is
# then plotted in a tilted bar plot.
top_10_occupations <- loan_data %>% 
  group_by(Occupation) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count)) %>% 
  top_n(10, Count)
ggplot(top_10_occupations) +
  geom_col(aes(reorder(Occupation, Count), Count)) +
  labs(x = "Occupation ordered by Count") +
  coord_flip()

# Create a dataframe with counts for each different type of loan status. 
loan_statuses <- loan_data %>% 
  group_by(LoanStatus) %>% 
  summarise(Count = n())
ggplot(loan_statuses) +
  geom_col(aes(LoanStatus, Count)) +
  coord_flip()

# Create and plot a dataframe with the counts for all the exsisting income
# ranges in the data set. The labels are tilted by 20 degrees for readability.
count_income_ranges <- loan_data %>% 
  group_by(IncomeRange) %>% 
  summarise(Count = n())
ggplot(count_income_ranges) +
  geom_col(aes(IncomeRange, Count)) +
  theme(axis.text.x=element_text(angle=20, vjust=0.5)) +
  coord_flip()

# Create and plot a dataframe with the counts for all exsisting listing
# categories in the data set.
ranked_listing_categories <- loan_data %>% 
  group_by(ListingCategory) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))
ggplot(ranked_listing_categories) +
  geom_col(aes(reorder(ListingCategory, Count), Count)) +
  labs(x = "Listing Categories ordered by count") +
  coord_flip()

```

From the above plots we can see that most of the loan takers are employed but
the type of occupation is seldom given with the vague "Professional" and "Other"
occupation types both being in the top ten. A majority of loans are still being
repaid but there are also an substantial amount of past due, defaulted or
charged-off loans.

Further, the income range looks to be fairly normalized with an expected value
somewhere around $50,000. Lastly, we have the listing categories for the loan
listings and we can see that roughly half of the the reasons given for the loans
through prosper is debt consolidation followed by the once again rather vague
"Not Available" and "Other" categories in the top three.

Now, after have gone through and had an initial look at all the variables in the
data set let's revisit and plot some of the more interesting numerical variables
to see how they are distributed over time.

```{r echo=FALSE, warning=FALSE, fig.align="center", Univariate_Plots_3}

# Bin plots for loan origination and closing dates. Binned at 1 month periods
# for 100 months for the period where we have available data for the whole
# month. We also calculate the cumulative sum of Originations and Closing per
# entry to get the amount of active listings for the period 2005-12-01 to 2014-03-01.
ggplot(subset(loan_data, 
              LoanOriginationDate < as.Date("2014-03-01") & 
                LoanOriginationDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(LoanOriginationDate), stat = "bin", bins = 100) +
  scale_y_continuous(limits = c(0, 6000),
                     breaks = seq(0, 6000, 1000))

ggplot(subset(loan_data, 
              ClosedDate < as.Date("2014-03-01") & 
                ClosedDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(ClosedDate), stat = "bin", bins = 100) +
  scale_y_continuous(limits = c(0, 6000),
                     breaks = seq(0, 6000, 1000))

Originations <- loan_data %>%
  group_by(LoanOriginationDate) %>%
  summarise(OriginationCount = n(), 
            MeanBorrowerRate = mean(BorrowerRate))
Closings <- loan_data %>% 
  group_by(ClosedDate) %>% 
  summarise(ClosedCount = -n())
Delta <- tibble(Date = seq(ymd_hms("2005-11-15 00:00:00"), 
                           ymd_hms("2014-03-12 00:00:00"), 
                           "days")) %>% 
  left_join(Originations, 
            by = c("Date" = "LoanOriginationDate")) %>% 
  left_join(Closings,
            by = c("Date" = "ClosedDate")) %>% 
  rowwise() %>% 
  mutate(DeltaCount = sum(OriginationCount,
                           ClosedCount,
                           na.rm = TRUE))
ActiveListings <- Delta %>% 
  transform(Count = cumsum(DeltaCount))
ggplot(ActiveListings) + 
  geom_line(aes(Date, Count)) 

# Calculating and plotting monthly active listings changes over time
MonthlyActiveListings <- ActiveListings %>% 
  group_by(Month = floor_date(Date, "month")) %>% 
  summarise(Delta = sum(DeltaCount))
ggplot(MonthlyActiveListings) + 
  geom_line(aes(Month, Delta))

# Plotting mean borrower rates over time for originating loans
MonthlyOriginations <- Originations %>% 
  group_by(Month = floor_date(LoanOriginationDate, "month")) %>% 
  summarise(MonthlyMeanBorrowerRate = mean(MeanBorrowerRate))
ggplot(MonthlyOriginations) +
  geom_line(aes(Month, MonthlyMeanBorrowerRate))

# Plotting original loan amounts binned by $1000
ggplot(loan_data) +
  geom_histogram(aes(LoanOriginalAmount), bins = 35) +
  scale_x_continuous(breaks = seq(0, 35000, 5000))

# Loan term lenghts 
ggplot(loan_data) +
  geom_histogram(aes(Term), bins = 5) +
  scale_x_continuous(breaks = seq(0, 60, 12))

# Monthly Loan Payments
ggplot(loan_data) +
  geom_histogram(aes(MonthlyLoanPayment), bins = 46) +
  scale_x_continuous(breaks = seq(0, 2300, 100)) +
  theme(axis.text.x=element_text(angle=20, vjust=0.5))

# Borrower rates histogram
ggplot(loan_data) +
  geom_histogram(aes(BorrowerRate), bins = 51) +
  scale_x_continuous(breaks = seq(0, 0.5, 0.1))

# $DebtToIncomeRatio using log10 scale for y to compensate for the number of
# values around 0.25
ggplot(loan_data) +
  geom_histogram(aes(DebtToIncomeRatio), bins = 41) +
  scale_x_continuous(breaks = seq(0, 10, 0.5)) +
  scale_y_log10()

```

By plotting the above values we discovered some interesting facts such as that
the loan term probably are locked at one, three or five years. We also saw that
the usual amounts being borrowed are grouped around even $5000 numbers with a
maximum at $35000. Lastly, the loan origination dates clearly shows effects from
the 2008 recession and also an peak in new loans later years which we still
aren't able to explain. 

Let's proceed with going over our findings once more before investigating
inter-variable relations in the bivariate section.

# Univariate Analysis

### What is the structure of your dataset?

The original data set contains more variables than what we could practically
cover in one go, so to narrow them down we posed some questions about the data.
Let's revisit these questions to see if we made any discoveries worth noting
already.

First, to see who is using the service we can look at the following variables
mentioned:

* Employment Status
* Occupations
* Income Ranges
* IsBorrowerHomeowner
* etc.

Based on the summary data and plots presented, our typical loan taker is
employed, with an unspecified occupation, and probably an income of $25,000 to
$50,000. He/she is currently an homeowner and have debts of about a ratio of
0.22 of their income.

The reason for the loan is most likely debt consolidation with home improvement
and business lying as distant seconds among the specified reasons as seen in the
histogram with ranked listing categories.

To see what eventually happened with the loan we can have a look at the loan
status bin plot giving an overview over the different statuses for all the loans
in the data set. Out of a little over 100,000 loan listings we have a little
over 10,000 that have been defaulted or charged-off(> 150 days overdue with no
reasonable expectation of sufficient payment).

### What is/are the main feature(s) of interest in your dataset?
### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Loan origination dates vs. closed dates together with the data on overdue,
defaulted, or charged-off loans. Using these features together with the above
calculated active loans variable it would be interesting to dive deeper and see
how, and when, the 2008 recession affected the loans taken. 

Another features worth looking in to is the borrower rates which mostly pikes my
interest due to the unclear form of the distribution. Investigating what
variables are correlated and how they affect this blob of values centered
somewhere around 0.2 would be very interesting and perhaps a good candidate for
analysis by creating a regression model.

### Did you create any new variables from existing variables in the dataset?

Using the Loan origination dates together with the closing dates I calculated a
new variable called active listings to show the volume of current loans on the
service. The calculation where made by taking the difference between originated
and closed loans for each date during a period between 2005 and 2014 and the
calculate the cumulative sum of those differences.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

In addition to make sure that all variables were read in using the correct data
type, I also choose how to handle NA values. For NA values for nominal data was
substituted with an preexisting category being best suited to improve the
readability of following histograms. Numeral values where set to 0 for ordinal
variables where NAs where present this in order to ease calculations in the
analysis while not influencing other statistical measures.

# Bivariate Plots Section

To continue follow our interesting findings regarding the effects of the 2008
recession on active loan volumes let's plot the change in loan statuses and 
listing categories for the period.

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_1}

# Loan closed statuses for loans over time.
ggplot(subset(loan_data, 
              ClosedDate < as.Date("2014-03-01") & 
                ClosedDate > as.Date("2005-12-01"))) +
  geom_histogram(aes(ClosedDate, fill = LoanStatus), 
                 stat = "bin",
                 bins = 100,
                 position = "fill") +
  scale_fill_manual(values = c("#c7e9c0",
                               "#74c476", 
                               "#238b45",
                               "#005a32"))

# Ratio for listing categories over time.
ggplot(loan_data) +
  geom_histogram(aes(ListingCreationDate, fill = ListingCategory),
                 bins = 100, 
                 position = "fill") +
  scale_fill_manual(values = c("#edf8e9",
                               "#c7e9c0",
                               "#a1d99b",
                               "#74c476", 
                               "#41ab5d",
                               "#238b45",
                               "#005a32",
                               "#eff3ff",
                               "#c6dbef",
                               "#9ecae1",
                               "#6baed6",
                               "#4292c6",
                               "#2171b5",
                               "#084594",
                               "#feedde",
                               "#fdd0a2",
                               "#fdae6b",
                               "#fd8d3c",
                               "#f16913",
                               "#d94801",
                               "#8c2d04"))

```

First, we can see that there have been some change in the gathering of the data
in the shift between 2007 and 2008. Looking at loan statuses we can see that the 
ratio for Chargedoff and Defaulted loans did shift sharply with the amount of 
Chargedoff loans skyrocketing. Analogously, the listing categories went from 
mostly Not available to a mix of all different categories at the same period.

Looking at some of the numerical values in our data set we unfortunately cannot 
see any meaningful correlation between variables, except for the connection 
between monthly payments and the original amount of the loan.

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_2}

# Plotting correlation matrix
ggpairs(subset(loan_data, 
               select = c("BorrowerRate",
                          "LoanOriginalAmount",
                          "MonthlyLoanPayment",
                          "DebtToIncomeRatio",
                          "Investors")),
        lower = list(continuous = wrap("points", 
                                       size = 0.2,
                                       alpha = 0.1))) +
    theme(axis.text.x=element_text(angle=20, vjust=0.5))


```

Following our slight drawback from analyzing the continuous variables we instead 
focus on the categorical ones. Here we can start to see some slight trends in 
the relationships between all our supporting variables and the borrower rate, 
however not one single variable seems significant enough to draw any conclusions. 

```{r echo=FALSE, warning=FALSE, fig.align="center", Bivariate_Plots_3}

# Plotting the term(cut into categories) vs given borrower rate
ggplot(loan_data) +
  geom_boxplot(aes(cut(Term, c(0, 12, 36, 60)),
                 BorrowerRate)) +
  coord_flip()

# Plotting IncomeRange vs borrower rate
ggplot(loan_data) +
  geom_boxplot(aes(IncomeRange,
                 BorrowerRate)) +
  coord_flip()

# Plotting ListingCategory vs borrower rate (Ranked by BorrowerRate median)
ranked_listing_categories_by_borrower_rate <- loan_data %>% 
  group_by(ListingCategory) %>% 
  summarise(Median = median(BorrowerRate)) %>% 
  arrange(Median)
loan_data$ListingCategory <- 
  factor(loan_data$ListingCategory,
         levels = ranked_listing_categories_by_borrower_rate$ListingCategory)
ggplot(loan_data) +
  geom_boxplot(aes(ListingCategory, BorrowerRate)) +
  coord_flip()

# Plotting Occupation vs Borrower rate (Ranked by BorrowerRate median)
ranked_occupation_by_borrower_rate <- loan_data %>% 
  group_by(Occupation) %>% 
  summarise(Median = median(BorrowerRate)) %>% 
  arrange(Median)
loan_data$Occupation <- 
  factor(loan_data$Occupation,
         levels = ranked_occupation_by_borrower_rate$Occupation)
ggplot(loan_data) +
  geom_boxplot(aes(Occupation,
                 BorrowerRate)) +
  coord_flip()

# Plotting HomeOwner vs BorrowerRate
ggplot(loan_data) +
  geom_boxplot(aes(IsBorrowerHomeowner,
                 BorrowerRate)) +
  coord_flip()

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?

We first had a look at the ratios for loan statuses and listing categories to 
see if we could see any effects from the 2008 depression in our data other than 
the decrease in the volume of loans. In the loan status data we could see an 
increase in the part of loans being defaulted or Chargedoff followed by a 
decrease in ratio when the economy recovered. For the listing categories we saw 
that the biggest category has been and continues to be debt consolidation, 
except for the obvious gaps in the data during 2008 there doesn't seem to have 
been any greater impacts on the types of loans taken through prosper during or 
after the time of 2008.

Turning our focus towards the borrower rate for loans I split up the 
visualizations into two steps depending on if the supporting variables were of 
a categorical or continuous type. First looking at the continuous variables we 
can see that the original amount of the loan, together with the size of the 
monthly payments and number of loan givers affect the borrower positively when 
increasing, there is a negative correlation meaning that an increase in either 
variable is connected with an decrease in the borrower rate for the loan.
The only variable we focus on that increases the borrower rate is the debt to 
income ratio which shows an slightly positive correlation.

It is worth mentioning that none of the above relations did show any stronger 
correlation all having a ratio below 0.5.

For the categorical variables all of them pointed to correlations between the 
borrower rate and income, home ownership, term, occupation, and category of the 
loan. It would definitely be interesting to pursue the analysis of the 
correlation between borrower rates and these different variables further in the 
following multivariate section.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
### What was the strongest relationship you found?

The only really strong relationship I could find in the above analysis were the 
relation between the original amount of the loan and monthly payments. This does 
seem a bit trivial, since you could come to this conclusion without any 
statistical analysis at all but it could be worth mentioning.

# Multivariate Plots Section

First we take a look at the loan original amount and borrower rate in relation 
to income range, term, debt to income ratio, and if the borrower is a homeowner 
or not.

```{r echo=FALSE, warning=FALSE, fig.align="center", Multivariate_Plots_1}

# Borrower rate density by IncomeRange
ggplot(loan_data) +
  geom_density(aes(BorrowerRate, color = IncomeRange))

# Borrower rate density by loan term 
ggplot(loan_data) +
  geom_density(aes(BorrowerRate, color = cut(Term, c(0, 12, 36, 60))))

# Borrower rate density by DebtToIncomeRatio
ggplot(loan_data) +
  geom_density(aes(BorrowerRate, 
                   color = cut(DebtToIncomeRatio,
                               c(0, 
                                 0.14, 
                                 0.22, 
                                 0.32, 
                                 10.010)))) +
  labs(color = "DebtToIncomeRatio")

# Borrower rate density for homeowner vs non-homeowner
ggplot(loan_data) +
  geom_density(aes(BorrowerRate, color = IsBorrowerHomeowner))

```

Next we have a look at Borrower rates vs loan original amount with color coding
for some different secondary variables.

```{r echo=FALSE, warning=FALSE, fig.align="center", Multivariate_Plots_2}

# Plotting BorrowerRate vs Loan original amounts colored by IncomeRange
ggplot(loan_data) +
  geom_point(aes(BorrowerRate, LoanOriginalAmount, color = IncomeRange),
             size = 0.5,
             alpha = 0.2) +
  scale_color_brewer(palette = "Greens")

# Plotting BorrowerRate vs Loan original amounts colored by LoanOriginationDate
ggplot(loan_data) +
  geom_point(aes(BorrowerRate, 
                 LoanOriginalAmount, 
                 color = cut(LoanOriginationDate, 
                             breaks = c(unique(floor_date(LoanOriginationDate, 
                                                   "year"))))),
             size = 0.5,
             alpha = 0.2) +
  labs(color = "LoanOriginationDate") +
  scale_color_brewer(palette = "Blues",
                     labels = c("2005",
                                "2006",
                                "2007",
                                "2008",
                                "2009",
                                "2010",
                                "2011",
                                "2012",
                                "2013"))

# Plotting BorrowerRate vs Loan original amounts colored by DebtToIncomeRatio
ggplot(loan_data) +
  geom_point(aes(BorrowerRate, LoanOriginalAmount, color = DebtToIncomeRatio),
             size = 0.5,
             alpha = 0.2) +
  scale_color_gradientn(colors = c("#fee5d9", 
                                   "#fcae91", 
                                   "#fb6a4a", 
                                   "#de2d26",
                                   "#a50f15"),
                        values = c(0, 
                                   rescale(0.14, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   rescale(0.22, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   rescale(0.32, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   1))

# Plotting BorrowerRate vs Loan original amounts colored by Term
ggplot(loan_data) +
  geom_point(aes(BorrowerRate, 
                 LoanOriginalAmount, 
                 color = cut(Term, c(0, 12, 36, 60))),
             size = 0.5,
             alpha = 0.2) +
  scale_color_brewer(palette = "Oranges")
                   
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

From the group of scatter plots in this section we can see that the borrower 
rate for a given loan amount increases together with an longer term loan and
also with the borrower's debt to income ratio. We can see this since for any
given value of loan amount the color is getting darker the further to the right
we move. The inverse is also true for the borrower income range which shows that 
the borrower rate is generally lower for higher income borrowers given the same
loan amount. Lastly, we have the year of origin which does not show any similar
patterns, but instead there is a step change in the minimum loan amount and
maximum borrower rate. This is most likely due to a policy change at prosper
taking place somewhere 2007-2008.

Moving on to the density plots we can see some interesting observations in the 
plot comparing borrower rate and income ranges. The not employed 
status is in general generating the worst borrower rate but surprisingly, not
displaying anything at all is better and comparable to the lowest income range
bracket. Something even more surprising is that $0 income shares roughly the same
shape as income ranges from $50000 and upwards. 

Looking at the borrower rate density for loan terms we can see that shorter
term loans do get lower borrower rates but that there is a turning point
somewhere after 36 months where the uncertainty from a longer term length take
the overhand over other factors deciding the borrower rate. This can be seen in
the lower variance for loans with terms up to 60 months.

The debt to income ratio's density is neatly sorted in an expected order with
unreported and high ratios having peaks around 0.3 in borrower rates and lower
rates having peaks between 0.2 and 0.1. The only thing standing out is that the
fourth quartile (0.32, 10] almost has two equally high peaks at both 0.2 and
0.3. I believe this might be stemming from the fact that this span is the
longest of the four quartiles. Most likely the distribution for the lower and
upper half of this quartile has different distributions. (The fact that values
over 10 is all clumped together at this value is also definitely a factor.)

The last graph, showing the rates divided by home owner or not has start out 
with as we might expect with homeowner density being higher at lower interest
rate than non homeowner later switching to non homeowners being on top for
borrower rates just before hitting the 0.2 rate mark. Interesting to see
however was that at higher rates, 0.3 and above, being a homeowner or not does
not seem to affect the rate at all. Perhaps loan-takers given these kind of rates
has some other circumstances overshadowing the potential extra security being a
homeowner could bring.

### Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

My initial plan was to try create a linear model to predict borrower rates based
on the variables I had choose from the data set but going through the analysis up
until now I do not feel that I have been able to find enough data to make a
decision regarding what would constitute a suitable model. Rather than trying to
do anything based of loosely based assumption I will leave modelling for another
time.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your data set.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.



### Plot One
```{r echo=FALSE, warning=FALSE, fig.align="center", Plot_One}

ggplot(MonthlyActiveListings) + 
  geom_line(aes(Month, Delta)) +
  labs(title = "Monthly deltas in active loan listings") +
  theme_minimal()

```

### Description One

The first plot depicts the monthly deltas for active loan listings. The reason I
choose this is due to how clear the effect from the 2008 recession can be seen.
Not only did the increase of loans take a deep dive in the early half of 2008
but the number of loans actually decreased pretty much all the way into 2011. I
don't think any deeper conclusions can be drawn from this graph but it might be
a good observation of how the market has looked like coming into the 2010's.

### Plot Two
```{r echo=FALSE, warning=FALSE, fig.align="center", Plot_Two}

ggplot(loan_data) +
  geom_density(aes(BorrowerRate, color = cut(Term, c(0, 12, 36, 60)))) +
  labs(title = "Borrower rate density by term length",
       y = "Density",
       x = "Borrower rate",
       color = "Term") + 
  scale_color_hue(labels = c("12 months", "36 months", "60 months")) +
  theme_minimal()

```

### Description Two

The second graph I choose was the density curve for borrower rates colored by
term length. The reason for this is how the figure expresses the effect future
insecurity has on decisions today. The variance is high for both 12 and 36 month
loans but when the term increases further up to 60 months it shrinks. It is as
if the insecurity of the future makes the borrowers desicion for the borrower
rate more uniform since there is less individual data they can count with.

I can also add that unlike the other two graphs, which I can image that I could
have come up with even before taking the Udacity EDA course, this graph was
something new I hadn't seen, or perhaps paid attention to, during my years at
university.

### Plot Three
```{r echo=FALSE, warning=FALSE, fig.align="center", fig.width=12, fig.height=7, Plot_Three}

income_plot <- ggplot(loan_data) +
  geom_point(aes(BorrowerRate, LoanOriginalAmount, color = IncomeRange),
             size = 0.5,
             alpha = 0.2) +
  scale_color_brewer(palette = "Greens") +
  labs(y = "Loan original amount",
       x = "Borrower rate",
       color = "Income range") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_minimal()

origination_plot <- ggplot(loan_data) +
  geom_point(aes(BorrowerRate, 
                 LoanOriginalAmount, 
                 color = cut(LoanOriginationDate, 
                             breaks = c(unique(floor_date(LoanOriginationDate, 
                                                   "year"))))),
             size = 0.5,
             alpha = 0.2) +
  labs(color = "LoanOriginationDate") +
  scale_color_brewer(palette = "Blues",
                     labels = c("2005",
                                "2006",
                                "2007",
                                "2008",
                                "2009",
                                "2010",
                                "2011",
                                "2012",
                                "2013")) +
  labs(y = "Loan original amount",
       x = "Borrower rate",
       color = "Origination date") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_minimal()

debt_plot <- ggplot(loan_data) +
  geom_point(aes(BorrowerRate, LoanOriginalAmount, color = DebtToIncomeRatio),
             size = 0.5,
             alpha = 0.2) +
  scale_color_gradientn(colors = c("#fee5d9", 
                                   "#fcae91", 
                                   "#fb6a4a", 
                                   "#de2d26",
                                   "#a50f15"),
                        values = c(0, 
                                   rescale(0.14, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   rescale(0.22, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   rescale(0.32, 
                                           to = c(0, 1), 
                                           from = range(loan_data$DebtToIncomeRatio, 
                                                        na.rm = TRUE)),
                                   1)) +
  labs(y = "Loan original amount",
       x = "Borrower rate",
       color = "Debt to income ratio") +
  theme_minimal()

term_plot <- ggplot(loan_data) +
  geom_point(aes(BorrowerRate, 
                 LoanOriginalAmount, 
                 color = cut(Term, c(0, 12, 36, 60))),
             size = 0.5,
             alpha = 0.2) +
  scale_color_brewer(palette = "Oranges",
                     labels = c("12 months", "36 months", "60 months")) +
  labs(y = "Loan original amount",
       x = "Borrower rate",
       color = "Term") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_minimal()

grid.arrange(income_plot, origination_plot, debt_plot, term_plot, 
             ncol = 2,
             top = "Loan original amount vs borrower rate for different auxiliary features")

```

### Description Three

The third graph, and this might be cheating, is actually a combination of the four different graphs depicting borrower rates vs loan original amounts with different coloring for the distribution of income range, origination date, debt to income ratio, and term for the loans plotted. Even though there weren't any major breakthough due to this figure this time, it do show that by trying combination after combination of different variables patterns are eventually found and sometimes you stumble over something unexpected. (In this case the origination date graph.)

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the data set.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!